Large-scale deorphanization of Nematostella vectensis neuropeptide GPCRs supports the independent expansion of bilaterian and cnidarian peptidergic systems

Version 1

Abstract

Abstract

Neuropeptides are ancient signaling molecules in animals but only few peptide receptors are known outside bilaterians. Cnidarians possess a large number of G protein-coupled receptors (GPCRs) – the most common receptors of bilaterian neuropeptides – but most of these remain orphan with no known ligands. We searched for neuropeptides in the sea anemone Nematostella vectensis and created a library of 64 peptides derived from 33 precursors. In a large-scale pharmacological screen with these peptides and 161 N. vectensis GPCRs, we identified 31 receptors specifically activated by one of 14 peptides. Mapping GPCR and neuropeptide expression to single-cell sequencing data revealed how cnidarian tissues are extensively wired by multilayer peptidergic networks. Phylogenetic analysis identified no direct orthology to bilaterian peptidergic systems and supports the independent expansion of neuropeptide signaling in cnidarians from a few ancestral peptide-receptor pairs.

Introduction

The origin of neuropeptides predates the emergence of neurons and it is believed that these signaling molecules have been utilized in the most ancestral nervous systems (Jékely, 2021; Moroz et al., 2021; Yañez-Guerra et al., 2022). Neuropeptide-like molecules occur in all major animal clades with neurons, the Bilateria, Cnidaria and Ctenophora, and even in the neuron-less Placozoa and Porifera (Hayakawa et al., 2022; Koch and Grimmelikhuijzen, 2019; Nikitin, 2015; Sachkova et al., 2021; Yañez-Guerra et al., 2022). However, the deep relationships of animal neuropeptidergic systems, in particular between Bilateria and non-bilaterians, have remained elusive.

Most mature bioactive neuropeptides are 3-20 amino acids long and derive from longer proneuropeptide precursors through cleavage and other post-translational modifications. The same set of enzymes are involved in proneuropeptide processing across animals. These include prohormone convertases, which recognize dibasic cleavage sites flanking the active peptides, and peptidyl-glycine alpha-amidating monooxygenase (PAM), which convert C-terminal glycine residues to amide groups (Chufán et al., 2009; Seidah, 2011). Proneuropeptide precursors can contain a single peptide or multiple copies of identical or divergent peptide sequences separated by cleavage sites. The often repetitive structure and the presence of short active sequences interspersed with less-constrained interpeptide regions allow propeptide sequences to evolve relatively rapidly. Consequently, with increasing evolutionary distances it often becomes hard or impossible to recognize orthologous proneuropeptide sequences. Even within Bilateria, the orthology relationship of many neuropeptide families between protostomes and deuterostomes have only been recognized due to the orthology of their receptors (e.g., vertebrate orexin and insect allatotropin) (Alzugaray et al., 2019; Elphick et al., 2018; Mirabeau and Joly, 2013; Semmens et al., 2015).

Most neuropeptides signal through G protein-coupled receptors (GPCRs), which are larger seven-transmembrane proteins and show a slower evolutionary rate than proneuropeptide precursors. The cases where the evolution of both the proneuropeptides and their receptors could be reconstructed, revealed a strong pattern of coevolution between receptor and ligand (Elphick et al., 2018; Escudero Castelán et al., 2022; Grimmelikhuijzen and Hauser, 2012; Jékely, 2013; Mirabeau and Joly, 2013; Yañez-Guerra and Elphick, 2020). This coevolutionary pattern has been used to also trace the evolution of peptide families where the ligands are too divergent to retain phylogenetic signal. These analyses revealed more than 30 conserved peptidergic signaling systems across bilaterians (Elphick et al., 2018; Jékely, 2013; Mirabeau and Joly, 2013; Thiel et al., 2021; Yañez-Guerra et al., 2022).

The nervous system of cnidarians has long been known to be strongly peptidergic and proneuropeptides have been found across all major cnidarian clades (Koch et al., 2021; Koch and Grimmelikhuijzen, 2020, 2019). Most cnidarian neuropeptides are short amidated peptides, resembling those found in bilaterians. Genomic comparisons revealed three proneuropeptide families that were present in the cnidarian stem lineage. All three give rise to short amidated neuropeptides: GLWamides, GRFamides, and PRXamides (with “X” representing a variable amino acid residue) (Koch et al., 2021; Koch and Grimmelikhuijzen, 2020, 2019). Other neuropeptides are more specific to certain cnidarian groups such as QITRFamide and HIRamide to Hexacorallia or LRWamides to Anthozoa (Koch and Grimmelikhuijzen, 2020).

Peptides that are sufficiently similar to be considered orthologous between bilaterians and cnidarians are restricted to a few atypical neuropeptides. These include insulin-related peptides, glycoprotein-hormone-related peptides, trunk related peptides, nesfatin, and phoenixin (de Oliveira et al., 2019; Roch and Sherwood, 2014; Yañez-Guerra et al., 2022). The more common short amidated peptides, however, have no clear orthologs in bilaterians. Similarity is at most restricted to one or two C-terminal residues, such as between RFamide or Wamide peptides (Walker et al., 2009; Jékely, 2013; Williams, 2020). Sometimes this has been interpreted as evidence of common origin but there is no other evidence supporting potential orthologies for these cnidarian peptides and their bilaterian counterparts. Receptors for most cnidarian neuropeptides are still unknown, with two exceptions. These include receptors for Hydra vulgaris RFamide peptides and a receptor for a PRXamide maturation inducing hormone (MIH) in the hydrozoan Clytia hemisphaerica (Artigas et al., 2020; Assmann et al., 2014). Hydra RFamide peptides activate heterotrimeric peptide-gated ion channels belonging to the DEG/ENaC family and distantly related to bilaterian RFamide- and Wamide-gated ion channels (Dandamudi et al., 2022; Elkhatib et al., 2022; Gründer et al., 2022a). The Clytia MIH receptor is a class A GPCR, which shows a many-to-many ortholog relationship to a range of bilaterian neuropeptide families, several of which end in RFamide (Artigas et al., 2020).

Cnidarian genomes can encode a large number of class A GPCRs. The genome of the sea anemone Nematostella vectensis for example contains over 1000 GPCR genes (Krishnan and Schiöth, 2015). Earlier analyses suggested direct orthologous relationships between several cnidarian and bilaterian neuropeptide GPCRs such as orexin/allatotropin, somatostatin or neuropeptide Y receptors, beside others (Alzugaray et al., 2013; Anctil, 2009; Krishnan and Schiöth, 2015). However, these results were either only based on BLAST similarity, or limited phylogenetic analyses. More recent and comprehensive phylogenies suggest that cnidarian neuropeptide GPCRs are more closely related to each other than to bilaterian neuropeptide GPCRs and only show many-to-many, or few-to-many orthology with bilaterian receptors (Artigas et al., 2020; Hauser et al., 2022; Thiel et al., 2018). However, a comprehensive phylogenetic analysis of cnidarian GPCRs is still lacking. This, together with the paucity of experimentally characterized receptors, leaves our understanding of the evolution of eumetazoan peptidergic systems fragmentary.

Here, in a large-scale bioinformatic analysis, we map the global sequence diversity of metazoan class A GPCRs to identify neuropeptide GPCR candidates in cnidarians. We then use mass spectrometry and bioinformatics to compile a library of predicted N. vectensis neuropeptides. In a combinatorial pharmacological ligand-receptor assay, we test our peptide library against selected N. vectensis GPCRs and identify 31 neuropeptide receptors. By phylogenetic analysis, we reconstruct ancestral cnidarian GPCR A systems and their relationship to bilaterian systems. Finally, we map proneuropeptide and GPCR expression to a single-cell RNAseq dataset to analyze tissue-level peptidergic signaling in N. vectensis. Our results reveal extensive peptidergic networks in N. vectensis and support the independent expansion of neuropeptide signaling in Cnidaria and Bilateria.

Results and Discussion

Creation of a Nematostella neuropeptide library

To obtain a comprehensive library of endogenous neuropeptides in N. vectensis, we extended the list of known proneuropeptides (Hayakawa et al., 2019; Koch and Grimmelikhuijzen, 2020) by mass spectrometry and bioinformatic screening (Figure 1A). We first screened a N. vectensis transcriptome collection for sequences encoding a signal peptide. This predicted secretome was filtered with regular expressions to detect sequences with the repetitive dibasic cleavage sites (K and R in any combination) and amidation sites. In addition, we used standard BLAST searches using known cnidarian neuropeptide precursors. In parallel, we analyzed methanolic extracts from N. vectensis larval, juvenile and adult tissue by mass spectroscopy (LC-MS/MS) (Figure 1–figure supplement 1). The LC-MS/MS analysis confirmed the existence of peptides predicted from various precursor sequences (Supplementary file 1, 2), including precursors that produce identical peptides, such as the 2 proFLRNamide or proLRWamide1 precursors (Figure 1C) (Koch and Grimmelikhuijzen, 2020).

Identification of Nematostella vectensis neuropeptides.

(A) Pipeline to identify neuropeptides and their receptors and to reconstruct the evolution of cnidarian peptidergic signaling. (B) Peptide sequence logos of N-terminal and C-terminal peptide cleavage sites based on peptides detected by LC-MS/MS. Cleavage occurs at the dashed lines. (C) N. vectensis neuropeptide precursor schemes of peptides for which we identified a receptor, with sequence logos of the encoded peptide(s) on the left and length of precursor on the right. a = amide.

MS confirmed the occurrence of N-terminal peptide cleavage at dibasic KR-x, RR-x and KK-x sites, with “x” indicating the N-terminal amino acid of the resulting active peptide (Figure 1B, Supplementary file 1). Cleavage at such dibasic sites is typical for bilaterian precursors (Southey et al., 2008; Veenstra, 2000). In Nematostella, the cleavage site is often followed by a-Q residue as the most N-terminal amino acid of the peptide (Figure 1B). Cleavage at dibasic sites, however, does not occur in all precursors. Our MS data showed that some peptides are processed by cleavage N-terminal to a-Q residue without any basic residue but the cleavage site is instead often accompanied by an acidic (D or E) residue 1 or 2 positions N-terminally of the cleavage site (Figure 1B, Supplementary file 1). Another motif includes two proline residues in position 2 and 3 from the N-terminus of the peptide, and in some cases a single proline in either the 2nd or 3rd position (Figure 1B, Supplementary file 1). These proline-related cleavage sites are also often accompanied by an acidic residue flanking the peptide N-terminal of the cleavage site. Such non-dibasic cleavage sites have previously been proposed for Cnidaria (Hayakawa et al., 2022, 2019; Koch and Grimmelikhuijzen, 2019). Some precursors, such as the one for HIRamide, also showed alternative N-terminal cleavage of the same peptide copy, resulting in different versions of the same peptide (Figure 1–figure supplement 1, Supplementary file 1), indicating stepwise cleavage or controlled degradation of peptides. C-terminal peptide cleavage occurs at dibasic x-KR, x-RR, x-KK and in some cases at alternative monobasic Rxx-R sites, with a second basic amino acid 3 positions N-terminal of the cleavage site (Figure 1B, Supplementary file 1). This is similar to the C-terminal cleavage found in bilaterian propeptides (Southey et al., 2008; Veenstra, 2000). In many neuropeptides, we could also confirm peptide alpha-amidation by the conversion of a C-terminal glycine residue to an amide group.

Based on the MS data, we extended our regular expressions representing cleavage sites and re-screened the predicted secretome. With our combined approach, we could identify novel neuropeptide precursors, verify the processing of known neuropeptide precursors and refine cleavage site predictions. We used this information to prepare a list of 33 N. vectensis proneuropeptides, excluding potential isoforms of the same precursor but including 2 potential paralogs of the FLRnamide, LRWamide1, and QITRFamide precursors (Supplementary file 2). Our screen complements the list of known neuropeptides in N. vectensis (Hayakawa et al., 2022, 2019; Koch and Grimmelikhuijzen, 2020) with 15 new neuropeptide precursors. However, our list did not contain the recently identified GGYamide, GTEamide and IVLamide peptides (Hayakawa et al., 2022) or bursicon- and insulin-like peptides.

We inspected all precursors individually and predicted signal peptides, cleavage sites and amidation sites (Figure 1C, Supplementary file 2). Based on our cleavage-site predictions we then compiled a library of 64 synthetic Nematostella neuropeptides, including different versions of peptides from the same precursors and alternatively cleaved peptides that differ in the length of their N-terminus (Supplementary file 2 and Supplementary file 3).

Analysis of metazoan class A GPCRs and selection of N. vectensis neuropeptide-receptor candidates

To identify neuropeptide-GPCR candidates in N. vectensis, we focused on class A GPCRs representing the main type of neuropeptide receptors in bilaterians (Jékely, 2013; Mirabeau and Joly, 2013). We first aimed to get an overview of the diversity of class A GPCRs across metazoans and screened transcriptomes of nine cnidarian, six bilaterian, two placozoan, three ctenophore and five sponge species for these receptors (Supplementary file 5, 6). The number of GPCRs across Metazoa varies by species and seems not to correlate with phylogenetic affiliation (Figure 2A). From the combined N. vectensis transcriptomes (see Supplementary file 5) we initially detected a total of 1061 class A GPCR sequences. We analyzed these together with over 7000 GPCRs retrieved from the other species. As a reference, we used the experimentally confirmed and annotated human, Drosophila melanogaster, Caenorhabditis elegans, and Platynereis dumerilii sequences. We carried out a sequence-similarity-based clustering analysis of all sequences. This revealed only a few well-connected clusters of receptors conserved across all major metazoan lineages (Figure 2–figure supplement 1). These include the leucine-rich-repeat containing (lrrc) GPCRs, which are receptors for larger peptides such as bursicons and related glycoprotein hormones or for relaxins and related insulin-like peptides. Other well-connected GPCR clusters were more restricted to the individual phylogenetic groups. Mammalian olfactory receptors represent such a GPCR cluster with limited taxonomic breadth (Figure 2–figure supplement 1).

Cluster map of selected class A GPCRs.

(A) Number of class A GPCRs identified by HMMer search in the different investigated species. (B) Relationship of species used for cluster analysis in C. (C) Cluster analysis of major class A GPCR groups from cnidarian, bilaterian and placozoan species. Each dot represents a GPCR sequence with color-coding and symbols according to the phylogeny in B. Connecting lines between single sequences show similarity with P-values indicated in the top right. Cluster annotations are based on deorphanized bilaterian class A GPCRs. Abbreviations in C: ACh = acetylcholine, lrrc = leucine rich repeat containing, P2Y = purinergic P2Y receptor. Silhouette images in B were taken from phylopic.org.

Figure 2–source data 1. Raw cluster analysis CLANS file.

For a more detailed analysis, we then reduced our sampled species to the cnidarian, the bilaterian with experimentally confirmed GPCRs and Petromyzon marinus, and the two placozoan species (Figure 2B). We re-run the clustering analysis and filtered out non-connected single sequences, species-specific cnidarian clusters and unconnected clusters only containing bilaterian or placozoan sequences. The two major clusters in our GPCR map contain bilaterian and cnidarian sequences, interspersed with only a few placozoan GPCRs (Figure 2C). In both major clusters, cnidarian and bilaterian sequences form separate subclusters, rather than cnidarian GPCRs being interspersed with bilaterian GPCRs. One of these clusters contains bilaterian GPCRs for low-molecular-weight neurotransmitters, including monoamines, trace amines, acetylcholine, melatonin and ATP. This cluster is weakly connected to the opsins. The second major cluster contains most bilaterian neuropeptide receptors (Figure 2C - neuropeptide II, III, gamma) as well as the bilaterian-specific chemokine, purinergic-P2Y and fatty acid receptors. There is a second cluster of bilaterian neuropeptide receptors weakly connected to this main cluster (Figure 2C - neuropeptide I), which has only few cnidarian sequences connected to it. Based on this cluster analysis, we selected 161 Nematostella GPCR sequences. We focused on full-length GPCRs that are associated with the bilaterian neuropeptide GPCR clusters. In addition, we chose candidates from non-connected clusters that are uncharacterized but ancestral to all cnidarians, except the leucine-rich repeat-containing GPCRs, which are known to be activated by insulin-related and glycoprotein-hormone-related peptides in bilaterians.

Deorphanization of 31 Nematostella neuropeptide receptors

To experimentally identify Nematostella neuropeptide GPCRs, we tested our selection of 161 N. vectensis GPCRs in a pharmacological assay for activation by peptides from our peptide library (Figure 3A, Supplementary file 3, 8). We expressed the candidate GPCRs in mammalian cells, together with a promiscuous Gqi protein and a luminescent G5A reporter (Figure 3A, Supplementary file 7). The peptides were separated into different mixes (Supplementary file 3), which were then tested on each GPCR (Supplementary file 4). Receptor-mix combinations that gave a positive signal were further resolved by testing the individual peptides of the mix to identify the activating ligand (Supplementary file 4). Peptides activating a receptor were then tested at different concentrations to record dose-response curves and determine EC50 values for each peptide-receptor pair (Supplementary file 9).

Dose-response curves of Nematostella neuropeptide GPCR pairs.

(A) Pharmacological assay and pipeline to identify peptide-GPCR pairs. (B) Dose-response curves of peptide-GPCR pairs with log peptide concentration plotted against normalized luminescence. GPCRs that are activated by the same peptide(s) are grouped together with peptide sequence shown above and peptide name highlighted in black. If several peptides activate the same receptor, peptide sequences are shown within the graph. Receptor identification number is encircled in the upper left of each curve, EC50 values are indicated in the lower right. (C) Histogram of EC50 values of peptide-GPCR pairs, showing only the lowest EC50 per GPCR. (D) Peptide-receptor pairings showing number of receptors activated by the different peptides. Connection strength indicates EC50 values.

Figure 3–source data 1. Tibble with all data points used to calculate the dose-response curves and EC50 values, in.csv format.

In this screen, we identified 31 N. vectensis GPCRs activated mostly in the nanomolar range by peptides from 14 different precursors (Figure 1C, 3B,C,D Supplementary file 10). The neuropeptides GLWL, PFHamide, VRHamide and QWamide each activate a different, single receptor (GLWL/R18 EC50 = 9E-7, PFHa/R36 EC50 = 1.2E-8, VRHa/R186 EC50 = 2.7E-10, QWa/R69 EC50 = 9E-7). Peptides ending in RPamide from the HRPamide and FRPamide precursors activate the same, single receptor R248. The HRPa peptide has a lower EC50 value, the FRPa peptide, however, has a lower threshold activation while reaching its maximum activation at higher concentrations with an overall more shallow curve slope (HRPa/R248 EC50 = 8.9E-9, HRPa/R248 slope = 1.75, FRPa/R248 EC50 = 2.3E-7, FRPa/R248 slope = 0.42) (Supplementary file 10).

The RFamide peptides with the sequence QGRFamide and QITRFamide are each encoded on separate precursors and despite their strong sequence similarities each activate separate receptors (QGRFa/R70 EC50 = 2.1E-6, QGRFa/R234 EC50 = 4.7E-9, QITRFa/R196 EC50 = 7.7E-11) although one of the two QGRFamide receptors is also activated by higher concentrations of QITRFamide (QITRFa/R70 EC50 = 1.7E-5). Given that R70 has EC50 values in a lower micromolar range for both RFamides, it is possible that this receptor has another unknown ligand that is more specific but shares some structural similarity to these RFamides. Two receptors are activated by HIRamide (R21 EC50 = 2.1E-9, R29 EC50 = 1.9E-7). The more sensitive R21, however, showed a strong base-activation at all tested concentrations, leading to a shifted EC50 for which we adjusted the minimum values (Figure 3–figure supplement 1). We also tested different copy versions of HIRamide peptides and most activated the two receptors in a similar concentration range (Figure 3–figure supplement 1).

Four receptors are activated by FLRNamide (FLRNa/R26 EC50 = 7.4E-8, FLRNa/R187 EC50 = 7E-8, FLRNa/R197 EC50 = 1.1E-8, FLRNa/R230 EC50 = 4.7E-9), and one of these is also sensitive to higher concentrations of FYRVamide (FYRVa/R197 EC50 = 1.1E-6). However, we only tested the non-modified FLRNamide peptide and not the phenylacetyl-LRNamide as described to exist in sea anemones (Grimmelikhuijzen et al., 1990) which may be a better ligand.

The three LRWamide peptides QGLRWamide (LRWa1), QGGLRWamide (LRWa2) and GPPQGLRWamide (LRWa3), which are each encoded in a single copy on separate precursors (Figure 1C), cross-activate 4 different GPCRs. Each LRWamide seems to have one preferred receptor (LRWa1/R19 EC50 = 9.1E-8, LRWa2/R213 EC50 = 3.1E-8, LRWa3/R204 EC50 = 7.4E-10) plus a fourth GPCR (R193) which appears similarly sensitive to all LRWamide peptides (LRWa1/R193 EC50 = 6.6E-11, LRWa2/R193 EC50 = 2E-10, LRWa3/R193 EC50 = 6.9E-10). Receptor 19 and 204 have clearly shifted curves and lower EC50 values for their preferred peptide, while Receptor 213 has similar EC50 values for LRWa1 (EC50 = 5.4E-8) and LRWa2 (EC50 = 3.1E-8) but is more sensitive to LRWamide 2 at lower concentrations while reaching its maximum activation later.

Finally, the PRGamide peptide, which belongs to the ancestral cnidarian PRXamides, activates at least 13 different receptors (R28, R32, R198, R199, R200, R202, R210, R211, R219, R220, R221, R222, R223) in Nematostella with EC50 values between 1.6E-10 and 4.5E-8. We also tested R28 and R32 with longer versions of the PRGamide, as these were described elsewhere based on mass spectrometry data (Hayakawa et al., 2019). The higher EC50 values for these longer PRGamides (Figure 3–figure supplement 2) and the sequences on the precursor themselves (Supplementary file 2), however, suggest that these are not fully processed and the actual PRGamide is a tetrapeptide (GPRGamide and APRGamide) as previously suggested (Koch and Grimmelikhuijzen, 2020). The same is likely true for the longer QGRFamide version (QGRFGREDQGRFamide) (Hayakawa et al., 2019) which is also likely not fully processed as in this case the activation of the QGRFamide receptors in the initial screen was much lower for the mix that contained the longer version than for the mix containing the fully processed QGRFamide peptide (Supplementary file 4 - mix 4 vs. mix 6).

At least nine neuropeptide GPCR families are ancestral to cnidarians

To reconstruct the evolution of neuropeptide receptors in animals, we analyzed the phylogenetic relationships of the 31 deorphanized N. vectensis neuropeptide GPCRs to other cnidarian and bilaterian GPCRs. From the clustermap (Figure 2), we chose sequences with connection to the bilaterian and cnidarian neuropeptide GPCR clusters and calculated phylogenetic trees. In an initial analysis, we found that the bilaterian chemokine, purino, fatty acid and other related bilaterian-specific receptors are likely a diverged ingroup of the bilaterian neuropeptide gamma rhodopsin receptors (Figure 4–figure supplement 1,2, Supplementary file 11). Gamma rhodopsin receptors are specific to bilaterians and include somatostatin/allatostatin A, opioid/somatostatin/allatostatin C, kisspeptin and melanin-concentrating hormone receptors (Mirabeau and Joly, 2013; Thiel et al., 2021). We then deleted redundant group-specific expansions of loosely connected orphan clusters, the non-neuropeptide chemokine and related receptors, decreased the number of bilaterian species and calculated a detailed neuropeptide-GPCR phylogeny (Figure 4, Supplementary file 11).

Phylogeny of metazoan class A neuropeptide GPCRs.

(A) Phylogeny of species used in B. (B) Phylogeny of neuropeptide GPCRs with names of ligands. Branches are color coded according to A. Branches of deorphanized Nematostella GPCRs end in an asterisk. Alternating shades behind the tree branches highlight different monophyletic groups. Roman Numbers 1-3 and greek symbol gamma indicate approximate neuropeptide clusters shown in Figure 2. Left half circle of branch support indicates aBayes and the right half circle aLRT-SH-like support values. Detailed annotations in Supplementary file 11. (C) Table with number of receptors per group as highlighted in receptor phylogeny with a straight line indicating no receptor present. Two-letter abbreviations on top correspond to species in A. Abbreviations: a = amide, B = Bilateria, CCK = cholecystokinin, GnRH = gonadotropin releasing hormone, MIH = maturation-inducing hormone, Nm-U = neuromedin U, NpFF = neuropeptide FF, NpY/F = neuropeptide Y/neuropeptide F, P = Placozoa, PrP = prolactin releasing peptide, R.# = Nematostella GPCR number, sNpF = short neuropeptide F, t-FMRFa = trochozoan FMRFamide, TRH = thyrotropin releasing hormone.

Figure 4–source data 1. Raw sequences used for tree building,.fasta format.

Figure 4–source data 2. Aligned sequences used for tree building.

Figure 4–source data 3. Trimmed sequence alignment used for tree building.

Figure 4–source data 4. Tree file in nexus format.

The majority of known neuropeptide GPCRs are grouped into three main bilaterian clusters and two main cnidarian clusters (Figure 4B). Within the two cnidarian clusters there are at least nine neuropeptide GPCR families that are ancestral to Cnidaria, all of which are represented with clear anthozoan and medusozoan orthology groups. The deorphanized N. vectensis neuropeptide GPCRs belong to at least seven of these ancestral cnidarian families (Figure 4B, C).

One of the two main cnidarian clusters seems to have only expanded late within the Hexacorallia branch, with an especially high number of more than 100 GPCRs in N. vectensis (Figure 4B,C). The entire cluster is expanded in N. vectensis and Exaiptasia pallida, but neither expanded in Corallium rubrum, nor in medusozoan species. This cluster likely expanded from a single ancestral cnidarian receptor family as it only contains a single branch of medusozoan sequences. Alternatively, this cluster can be traced back to two peptidergic systems, with a loss of medusozoan sequences in one of them. This cluster contains the more promiscuous and less sensitive N. vectensis QGRFamide GPR70, the QWamide GPR69 and the PFHamide GPR36. The second major cnidarian neuropeptide GPCR cluster contains at least eight ancestral cnidarian receptor families, based on the presence of clear anthozoan and medusozoan orthology groups, with deorphanized N. vectensis receptors in six of them (Figure 4B,C). The QITRFamide GPR196 belongs to a small group of hexacorallian receptors with no clear orthologs in C. rubrum or any medusozoan species, which might either represent a separate ancestral group with no medusozoan representative sequences or a strongly diverged Hexacorallia subcluster. This is in accordance with the absence of QITRFamide peptides in Octocorallia and Medusozoa (Koch and Grimmelikhuijzen, 2020). The QGRFamide GPR234 belongs to an ancestral cnidarian GPCR family, which is in accordance with the existence of QGRFamide or GRFamide peptides across Cnidaria (Koch and Grimmelikhuijzen, 2020, 2019). This group is slightly expanded in the medusozoans Clytia hemisphaerica and Hydra vulgaris. The entire QGRFamide family seems related to the LRWa GPR19 containing group, which has otherwise no direct orthologous sequences in medusozoans. The GLWL peptide GPR18 belongs to an ancestral cnidarian family that shows a slight expansion in N. vectensis. The receptors for the peptides LRWamide, FLRNamide and HIRamide are closely related as part of an anthozoan/hexacorallian expansion, curiously twice in unrelated families. Each of the families is ancestral to cnidarians and one expanded in anthozoans and additionally contains the receptors for VRHamide and RPamides, while the other group only showed several sequences in the two analyzed Hexacorallia species N. vectensis and E. pallida. Both families contain receptors for the three peptides, LRWamide, FLRNamide and HIRamide, and both families only contain medusozoan sequences from R. esculentum and C. cruxmelitensis. The 13 PRGamide receptors of N. vectensis belong to two separate ancestral cnidarian families, both of which are expanded in anthozoans. One PRGamide family independently also expanded in Hydra vulgaris. The other PRGamide family contains the Clytia MIH receptor which is activated by Clytia PRXamides (Artigas et al., 2020), confirming a peptide-receptor coevolution in cnidarians for the ancestral PRXamides (Koch and Grimmelikhuijzen, 2019) and their receptor(s).

Cnidarian and bilaterian neuropeptide GPCR systems expanded after the cnidarian-bilaterian split from a few ancestral systems

Our phylogenetic analyses divided the cnidarian and the bilaterian neuropeptide receptors into very few major clusters, each containing either bilaterian or cnidarian sequences (Figure 4B). The relationship of these major clusters to each other is not well resolved. However, the general clustering (Figure 2C) in combination with the phylogenetic analysis (Figure 4B) shows that these groups represent many-to-many bilaterian-cnidarian orthology groups. Most of the bilaterian and cnidarian class-A neuropeptide GPCRs therefore diversified after the bilaterian-cnidarian split. The only family with an apparently consistent one-to-one orthology of an ancestral cnidarian and bilaterian branch of receptors is that of the bilaterian orphan receptor GPR19.

In addition there are a few branches with unclear phylogeny or restricted taxonomic breadth. A small group of anthozoan orphan receptors without medusozoan representatives shows affinity to the expanded cluster of bilaterian oxytocin/vasopressin, GnRH, achatin and neuropeptide S/CCAP receptors, suggesting a one-to-many orthology. This affinity is consistent in different trees (compare Figure 4 and Figure 4–figure supplement 1,2, or Supplementary file 11). Other orthology groups between Cnidaria and Bilateria are less clear and the potential branches in question are not stable between the different phylogenies. The bilaterian SIFamide GPCR grouped together with an orphan cnidarian branch, but varied in its position in other analyses (compare Figure 4 and Figure 4–figure supplement 1 and 2, or Supplementary file 11). This instability is also obvious due to the fact that the SIFamide GPCR is the protostome ortholog of the deuterostome NpFF GPCRs, which would have been expected to group together.

Accordingly, the NpFF receptor was also unstable in our analyses. The NpFF and SIFamide receptor group is usually well supported, but often shows long basal branches and these two together built in previous analyses a more separated branch with unstable relationship to other bilaterian receptor groups (Mirabeau and Joly, 2013; Thiel et al., 2021, 2018). The bilaterian QRFPR branch has a likewise unstable relationship to other bilaterian neuropeptide GPCRs in previous analyses but showed affinity to the SIFamide/NpFF GPCRs in some analyses (Jékely, 2013; Thiel et al., 2018). Accordingly, the affinity of some N. vectensis sequences to the QRFPR branch is also not stable in our different analyses (compare Figure 4 and Figure 4– figure supplement 1 and 2; Supplementary file 11). Two ancestral cnidarian branches grouped together with the Drosophila trissin receptor, but also this is an unstable grouping not present in our supplementary analyses (Figure 4–figure supplement 1 and 2). The trissin receptor is a generally peculiar bilaterian protostome receptor as no orthologous deuterostome sequences are known (Elphick et al., 2018; Mirabeau and Joly, 2013; Thiel et al., 2021). The orexin/allatotropin receptors showed no direct orthology to cnidarian receptors, but appear as a sister group to a branch with orphan bilaterian and placozoan GPCRs. Together, except for the GPR19 group, we could not find cnidarian receptors that show a consistent 1:1 orthology relationship to specific bilaterian receptors such as for example proposed for orexin/allatotropin, somatostatin, neuropeptide Y or tachykinin receptors (Alzugaray et al., 2019; Anctil, 2009; Krishnan and Schiöth, 2015). This is in accordance with other analyses that used a wider array of bilaterian neuropeptide GPCRs when comparing them to cnidarian GPCR sequences (Artigas et al., 2020; Hauser et al., 2022; Thiel et al., 2018) and found rather many-to-many orthologs, if any at all.

Cell-type-specific expression of neuropeptides and GPCRs and the peptidergic connectome of Nematostella

To analyze tissue-level peptidergic signaling in N. vectensis, we mapped the expression of neuropeptide precursors and the newly deorphanized neuropeptide GPCRs to a single-cell RNAseq dataset (Steger et al., 2022). The single-cell data are split into two sets, one consisting of pooled stages spanning 18 hours post-fertilization to 16-day-old primary polyp (developmental set) and the second set consisting of adult tissues only, similar to (Cole et al., 2023). The expression of GPCRs was in many cases low and only a small percentage of cells within a given cell type show receptor expression. We could not detect expression of GLWL receptor R18 and HIRamide receptor R29, consistent with the generally low expression of GPCRs in animals (Fredriksson and Schiöth, 2005; Regard et al., 2008; Soave et al., 2021; Sriram et al., 2019). In contrast, neuropeptide precursors are generally highly expressed in neurons (Smith et al., 2019) and we could detect all of them, except for the PFHamide that could not be mapped and was only present as a partial sequence in our combined transcriptome (Figure 1C). Most neuropeptide precursors show restricted expression in neuroglandular cells (Figure 5B), with the exception of the phoenixin and GLWL precursors (Figure 5–figure supplement 1). The broad expression of phoenixin suggests other functions for this molecule outside neuronal signaling, for example in mitochondrial regulation (Dennerlein et al., 2015; Yañez-Guerra et al., 2022).

Tissue-specific expression.

Dotplot for tissue-specific expression of peptide precursors and GPCRs. Red dots indicate expression in the developmental dataset, blue dots indicate expression in the adult dataset. Abbreviations: a = amide, e = embryonic, ect = ectoderm, endomes = endomesoderm, gland = glandular, muc = mucous, musc = muscle, neurogland = neuroglandular, PGC = primary germ cells, pharyng = pharyngeal, pSC = putative stem cells, R = receptor (GPCR), retrac = retractor.

Individual GPCRs were often restricted to a single or very few tissue types (Figure 5B for tissue type overview and Supplementary file 12 for cell-type-specific resolution). For peptides with multiple receptors, we often found distinct patterns of receptor expression. Individual PRGamide receptors are for example restricted to the retractor muscle (R221, R222), the gastrodermis (R211), embryonic putative stem cells (pSC) (R210) or a combination of neuroglandular cells and adult cnidocytes (R202) (Figure 5B). An exception is the PRGamide receptor R200, which shows expression in a wide array of tissues, including adult pSC, primary germ cells (PGCs), neuronal cells, immune cells, and embryonic ectodermal cells.

The different LRWamide receptors also signal to different tissue types. The receptor R204, which is most sensitive to LRWamide3, is strongest expressed in embryonic endomesodermal cells and the pharyngeal ectoderm, the receptor R19, which is most sensitive to LRWamide1 is found in neuroglandular cells, the receptor R213, which showed highest sensitivity to LRWa2 is found in neuroglandular and glandular mucousal cells and the receptor R193, which is similarly sensitive to all LRWamides is only present in glandular mucousal cells. The two receptors of the ancestral QGRFamide are also differentially expressed, with the receptor R70 only present in adult neuroglandular cells, while the highly specific R234 is additionally expressed in developmental and adult glandular mucousal cells and in some developmental retractor muscle cells.

Within the neuroglandular subset, each neuronal cell type expresses a unique combination of neuropeptide precursors and GPCRs (Figure 5–figure supplement 2 and 3; Supplementary file 12). Most peptide precursors are thereby restricted to only a few cell types. PRGamide for example, is restricted to two types of neuroglandular cells: the larval apical organ Nl(+)ao.PRGa cells and the Ne(-)PRDM14d cells, which persist in adults (Figure 5–figure supplement 2,3). This restricted expression of the PRGamide precursor is also in accordance with previously published in situ hybridizations that show a restricted expression in the larval apical organ (Gilbert et al., 2022). Many neuronal cell types only express a low number of different neuropeptide precursors, with few exceptions such as the mentioned Nl(+)ao.PRGa cells or the Nl(+)rxraB and N(+)GFI1B.prdm cells, which express between 4-8 different types of neuropeptide precursors, depending on the stage. The larval Nl(+)ao.PRGa cells express the PRGamide, VRHamide, OITRFamide, RPamide, LRE peptide, GLWL peptide and the phoenixin precursor, while only expressing two receptors: the PRGamide receptor R.200 with a lower average expression and the VRHamide receptor R186. The expression of both the PRGamide and VRHamide peptide precursors and receptors may indicate autocrine regulation in these cells. The second PRGamide-positive cell type Ne(-)PRDM14d that persists in adults, also shows in both datasets an expression of the PRGamide precursors and a different PRGamide receptor R.221, which is only found in few cell types and similarly may mediate autocrine feedback.

To determine the global organization of peptidergic signaling networks or the ‘peptidergic connectome’ (Bentley et al., 2016; Deng et al., 2019; Smith et al., 2019; Williams et al., 2017) in Nematostella, we constructed a multi-layered network for both the developmental and the adult dataset (Figure 6, Figure 6–figure supplement 1). In these networks, nodes represent cell/tissue types and links are defined from peptide-expressing cells to receptor-expressing cells taking into account expression levels and the EC50 values. Different peptide-receptor pairs (represented by different colors in Figure 6, Figure 6–figure supplement 1 B-D, F-L) form distinct layers in this multi-layer connectome. By modularity analysis, we subdivided the networks into three (developmental) or five (adult) modules, each dominated by peptides with multiple receptors: PRGamide, QGRFamide, FLRNamide and LRWamide (Figure 6–figure supplement 1 A,E). The cells at the center of these modules that are most strongly involved in peptidergic signaling are the Ne(+)pit1, Ns(+)otp, Nl(+)ao.PRGa, Nl(+)rxraB, and Ne(-)PRDM14d cells in the developmental dataset and the Ne(+)pit1, Ns(+)otp, Ne(+)otx.early and Ne(-)PRDM14d cells in the adult dataset.

Multi-layer peptidergic connectomes in Nematostella.

Peptidergic networks in the (A) developmental and (B) adult subset. Nodes represent cell types, connections represent potential peptidergic signaling from neuropeptide-expressing cells to cells expressing one or more of the receptors for that neuropeptide. Colors represent different peptide-receptor signal channels (the different layers in the multi-layer connectome).

Figure6–source data 1. Graph file of the multilayered peptidergic connectome in the developmental subset. A serialized binary file in tbl_graph format, to be analyzed in R.

Figure 6–source data 2. Graph file of the multilayered peptidergic connectome in the developmental subset. In Gephi.gexf format, to be analyzed in Gephi.

Figure 6–source data 3. Graph file of the multilayered peptidergic connectome in the adult subset. A serialized binary file in tbl_graph format, to be analyzed in R.

Figure 6–source data 4. Graph file of the multilayered peptidergic connectome in the adult subset. In Gephi.gexf format, to be analyzed in Gephi.

Discussion

We deorphanized 31 neuropeptide GPCRs of the sea anemone Nematostella vectensis and reconstructed their evolution and relationship to other cnidarian orphan GPCRs and bilaterian neuropeptide GPCRs. Our phylogeny suggests that the identified neuropeptide GPCRs belong to two major GPCR clades that expanded independently of each other. One clade diversified in stem Cnidaria into eight receptor families that are all present across medusozoans and anthozoans. The other clade diversified later within the anthozoans, likely from a single ancestral cnidarian receptor. The expansion of GPCR systems is a general feature of GPCR evolution and happened independently in different lineages and at different times during animal evolution (Beets et al., 2022; Mirabeau and Joly, 2013).

The cnidarian neuropeptide receptors we identified here show many-to-many orthology relationship to bilaterian neuropeptide GPCRs. We did not identify any direct receptor orthologs between cnidarians and bilaterians, indicating extensive parallel expansion of neuropeptide signaling in the two clades. Previously, some neuropeptides have been hypothesized to be orthologous across cnidarians and bilaterians, including neuropeptides ending in an RFamide (Jékely, 2013; Walker et al., 2009), which is a frequent motif (Elphick and Mirabeau, 2014; Jékely, 2013; Walker et al., 2009). We identified two N. vectensis RFamide receptors, one for the ancestral cnidarian neuropeptide QGRFamide (Koch et al., 2021; Koch and Grimmelikhuijzen, 2019) and one for the hexacorallian QITRFamide. These two N. vectensis receptors belong to either of the two major clades of cnidarian receptors, suggesting that new RFamide receptors and/or peptides can evolve independently. Likewise, within bilaterians there are many RFamides across protostomes and deuterostomes that are not orthologous to each other (Elphick and Mirabeau, 2014; Thiel et al., 2021). The cnidarian RFamide receptors are also more closely related to other cnidarian neuropeptide receptors than to bilaterian RFamide receptors, arguing against the orthology of cnidarian and bilaterian RFamides. C-terminal aromatic amino acids are found in various neuropeptides and also many monoamine receptors are activated by derivatives of aromatic amino acids (Bauknecht and Jékely, 2017; Elphick and Mirabeau, 2014). Aromatic amino acids might have been early ligands in the evolution of monoamine neurotransmission and ancestrally present as a common structural feature at the C-termini of neuropeptides activating peptide receptors.

Our phylogenetic analyses and GPCR resource will be useful to directly predict the ligand of neuropeptide receptors in other cnidarian species. We expect this because of the long-term coevolution of peptide-receptor pairs, that has been extensively documented in bilaterians. The orthology of six of the identified anthozoan N. vectensis PRGamide receptors to the PRXamide receptor of the hydrozoan C. hemisphaerica is the first example of a similar long-term ligand-receptor association across cnidarians. Receptors of ancient peptides such as PRXamides or GRFamide are therefore expected to be orthologous even in distantly related cnidarians. Other Hexacorallia-or Anthozoa-specific peptides such as HIRamide, FLRNamide and LRWamide (Koch and Grimmelikhuijzen, 2020) can be tested on directly orthologous receptors. Phylogenetic analyses combined with targeted deorphanization experiments of predicted orthologs could thus lead to the rapid characterisation of new receptors in other cnidarians. Our large-scale GPCR-peptide resource also allowed us to analyze tissue-level peptidergic networks in N. vectensis. With high-resolution single-cell resources becoming available in a larger number of species, it will be possible to predict the cellular targets of neuropeptide signals with increasing precision based on receptor expression. The unique combinations of proneuropeptides and neuropeptide receptors in the different neuronal cell types can also be used as markers to identify and characterize neuronal cell types. The uniquely specific combinatorial expression of proneuropeptides as neuronal markers parallels the situation found in bilaterian nervous systems, where proneuropeptides provide fingerprint-like identities to different neuronal cell types (Smith et al., 2019; Taylor et al., 2021; Williams et al., 2017). Finally, our work also opens up new avenues in experimental neuroscience in cnidarians (Bosch et al., 2017). With readily available genetic manipulation techniques in N. vectensis and other species (Ikmi et al., 2014; Nakanishi and Martindale, 2018; Paix et al., 2023) (Artigas et al., 2020; Sanders et al., 2018; Wittlieb et al., 2006), the identified neuropeptide-receptor interactions will enable genetic manipulations of both ligand and receptor(s), to reveal the biological functions of peptidergic signaling.

Overall, we identified receptors for peptides from nearly half of the Nematostella neuropeptide precursors, including the receptors of the ancient PRXamide and QGRFamide peptides. Future studies aiming at finding the remaining receptors could focus on GPCRs with no sequence similarity to known bilaterian neuropeptide GPCRs, including monoamine-related or leucine-rich-repeat receptors, or other types of receptors like DEG/ENaC-related ion channels (Gründer et al., 2022b; Mirabeau and Joly, 2013).

Supporting information

eLife Assessment

This work identifies cnidarian neuropeptides and pairs them to their GPCR, then shows that neuropeptide signaling systems have evolved and diversified independently in cnidarians and bilaterians. Neuropeptide-receptor partners were experimentally identified using established and widely used methodologies including single cell mapping, providing compelling evidence for the conclusions of the paper. This impressive accomplishment provides fundamental new insights into the evolution of neuropeptide signaling systems and will be of broad interest to neurobiologists and evolution of development researchers.

Joint Public Review:

Neuropeptide signaling is an important component of nervous systems, where neuropeptides typically act via G-protein coupled receptors (GPCRs) to regulate many physiological and behavioral processes. Neuropeptides and their cognate GPCRs have been extensively characterized in bilaterian animals, revealing that a core set of neuropeptide signaling systems originated in common ancestors of extant Bilateria. Neuropeptides have also been identified in cnidarians, which are a sister group to the Bilateria. However, the GPCRs that mediate the effects of neuropeptides in cnidarians have not been identified.

In this paper the authors perform a phylogenetic analysis of GPCRs in metazoans and report that the orthologs of bilaterian neuropeptide receptors are not found in cnidarians. This indicates that neuropeptide signaling systems have largely evolved independently in cnidarians and bilaterians. To accomplish this, they generated a library of putative and known neuropeptides computationally identified in the genome of the cnidarian sea anemone Nematostella vectensis. These peptides were systematically screened for their ability to activate any of the 161 putative Nematostella GPCRs.

This work identified 31 validated GPCRs. These, together with GPCRs that cluster with them, were then used to demonstrate the independent expansion of GPCRs in cnidarian and bilaterian lineages. The authors then mapped validated receptors and ligands to the Nematostella single cell data to provide an overview of the cell types expressing these signaling genes. In addition, the authors have begun to analyze neuropeptide signaling networks in N. vectensis by showing potential signaling connections between cell types expressing neuropeptides and cell types expressing cognate receptors.

This work is the most extensive pharmacological characterization of neuropeptide GPCRs in a cnidarian to date and thus represents an important accomplishment, and is one that will improve our understanding of how peptidergic signaling evolved in animals and its impact on evolution of nervous systems."

The reviewers did not detect any weaknesses in the work but ask that the authors comment on the following points:

1- It was not clear why the phylogenetic analysis included non-validated GPCRs that clustered with the validated peptidergic receptors. Would restricting the phylogenetic analyses only to confirmed peptidergic GPCRs alter the topology of the tree and subsequent conclusions of independent expansion?

2- Clearly, other neuropeptide signaling systems in cnidarians remain to be discovered but this paper represents a huge step forward.

3- There are limitations in what can be interpreted from single cell transcriptomic data but the data nevertheless provide the foundations for future studies involving i). detailed anatomical analysis of neuropeptide and neuropeptide receptor expression in N. vectensis using mRNA in situ hybridization and/or immunohistochemical methods and ii). functional analysis of the physiological/behavioral roles of neuropeptide signaling systems in N. vectensis

Author Response

The following is the authors’ response to the original reviews.

Reply to comments:

(1) It was not clear why the phylogenetic analysis included non-validated GPCRs that clustered with the validated peptidergic receptors. Would restricting the phylogenetic analyses only to confirmed peptidergic GPCRs alter the topology of the tree and subsequent conclusions of independent expansion?

Thank you for this comment. In general, phylogenetic analyses become more robust if a larger diversity and fuller complement of sequences are included. With very sparse sampling, sequences that are homologous but not orthologous may be misleadingly grouped together, because intermediate sequences have been left out. For tree building, we thus did not want to focus only on experimentally validated receptors but also on all receptors that are phylogenetically related to the validated receptors. Only this approach can ensure a comprehensive exploration of the relationship of peptidergic receptors. The broader phylogenetic approach was also essential to identify orthologs to the experimentally validated Nematostella receptors across other cnidarian species.

(2) Clearly, other neuropeptide signaling systems in cnidarians remain to be discovered but this paper represents a huge step forward.

We appreciate this assessment of the paper. We agree that many systems remain to be discovered. Our paper will also help with the identification of further receptors both in Nematostella as well as other cnidarian species. Please note that we have made specific receptor-ligand predictions for several cnidarian species based on our phylogenetic analysis. Our phylogenies could also help prioritize the study of the remaining orphan Nematostella GPCRs.

(3) There are limitations in what can be interpreted from single cell transcriptomic data but the data nevertheless provide the foundations for future studies involving i). detailed anatomical analysis of neuropeptide and neuropeptide receptor expression in N. vectensis using mRNA in situ hybridization and/or immunohistochemical methods and ii). functional analysis of the physiological/behavioral roles of neuropeptide signaling systems in N. vectensis

We fully agree with this comment. The analysis of the available single-cell sequence resources clearly represents only the first step of anatomical and functional analyses. Our aim was to place the identified peptide-receptor interactions into a whole-organism context with cell type resolution, to highlight the potential complexity of peptidergic signaling in this organism and to facilitate the exploration and conceptualisation of our biochemical screen.

Comments to authors

(1) In future, when preparing manuscripts, please use page and line numbers; it makes the task below for reviewers much easier!

We appreciate the suggestion and will do this for future manuscripts.

(2) In the abstract the term "extensively wired" is used. In the context of neuropeptide mediated volume transmission this may not be an appropriate term to use because use of the word "wired" is likely to be associated with point-to-point type classical synaptic transmission; "extensively connected" would be better.

Thank you for this comment. We have changed the text in the abstract to “extensively connected”.

(3) Introduction: Please change "seven-transmembrane proteins and show a slower evolutionary rate than proneuropeptide..." to "seven-transmembrane proteins that show a slower evolutionary rate than proneuropeptide..."

Changed.

(4) Under the section "Creation of a Nematostella neuropeptide library, what is meant by "our regular expressions"? This needs to be rephrased to make it clearer what is meant.

We have now rephrased the relevant sentence to make our approach clearer.

“This predicted secretome was filtered with regular expressions to detect sequences with the repetitive dibasic cleavage sites (K and R in any combination) and amidation sites, using a custom script from a previous publication (Thiel et al., 2021).”

and later:

“Based on the MS data, we included the additional, non-dibasic N-terminal cleavage sites into our script that uses regular expressions to search for repetitive cleavage sites (Thiel et al., 2024) and re-screened the predicted secretome.”

(5) Under the section "Creation of a Nematostella neuropeptide library" the phrase "differ in the length of their N-terminus" needs to be changed to "differ in the length of their N-terminal region". The N-terminus is, as its name implies, one end of the peptide/protein so it can't have a length as such.

Changed.

(6) Under the section "Analysis of metazoan class A GPCRs and selection of N. vectensis neuropeptide-receptor candidates",

Change:

"For a more detailed analysis, we then reduced our sampled species to the cnidarian, the bilaterian with experimentally confirmed GPCRs and Petromyzon marinus, and the two placozoan species (Figure 2B)."

To

"For a more detailed analysis, we then reduced our sampled species to cnidarians, bilaterians with experimentally confirmed GPCRs and Petromyzon marinus, and two placozoan species (Figure 2B)."

Changed.

(7) Under the section "Analysis of metazoan class A GPCRs and selection of N. vectensis neuropeptide-receptor candidates" - change "We re-run" to "We re-ran"

Changed.

(8) Throughout the paper reference is made to a variety of neuropeptides that have or are predicted to have an N-terminal pyroglutmate. However, these are referred to without indicating this post-translational modification e.g. QGRFamide.

This should be corrected throughout the paper, in the text, and figures. Two abbreviations for pyroglutamate are used in the literature:

pQ, which shows that the encoded amino acid is Q (Glutamine)

pE, which shows that the post-translationally modified amino-acid is glutamate (E)

In the neuropeptide field, pQ seems to be more widely used than pE, so our recommendation would be to use pQ.

In the revised version we now write pyroQ whenever we refer to the actual peptide. We now only use the peptide name without indicating this modification when we refer to the precursor of these peptides.

(9) The title for Figure 5 is rather short and vague. A title like "Tissue-specific expression of neuropeptide precursors and receptors in Nematostella" seems more appropriate

We appreciate the reviewer's input, and we have made the change accordingly. The revised figure legend now reads: “Tissue-specific expression of neuropeptide precursors and receptors (GPCRs) in N. vectensis.”

(10) All of the figures in the paper have been saved in bitmap format (e.g. tiff), which means that the resolution of the figures may end up being poor in the published article. All of the figures in this paper should be saved in vector format (e.g. eps) so that there is no loss of resolution when the size of the file/figure is reduced.

We have now uploaded all figures in vector format (.eps or .pdf) to prevent any loss of resolution.

(11) In Figure 3 - supplement 2 - the neuropeptides are referred to here as PRGamides and GPRGamides. Some consistency is needed here. And in Figure B, the G of one of the GPRGamides is not shown in black.

Thank you for spotting this mistake. We now give the correct peptide sequence in parenthesis as "GPRGamide". We also highlighted the missing GPRGamide in the figure.

Version 2

Abstract

Abstract

Neuropeptides are ancient signaling molecules in animals but only few peptide receptors are known outside bilaterians. Cnidarians possess a large number of G protein-coupled receptors (GPCRs) – the most common receptors of bilaterian neuropeptides – but most of these remain orphan with no known ligands. We searched for neuropeptides in the sea anemone Nematostella vectensis and created a library of 64 peptides derived from 33 precursors. In a large-scale pharmacological screen with these peptides and 161 N. vectensis GPCRs, we identified 31 receptors specifically activated by one of 14 peptides. Mapping GPCR and neuropeptide expression to single-cell sequencing data revealed how cnidarian tissues are extensively wired by multilayer peptidergic networks. Phylogenetic analysis identified no direct orthology to bilaterian peptidergic systems and supports the independent expansion of neuropeptide signaling in cnidarians from a few ancestral peptide-receptor pairs.

Introduction

The origin of neuropeptides predates the emergence of neurons and it is believed that these signaling molecules have been utilized in the most ancestral nervous systems (Jékely, 2021; Moroz et al., 2021; Yañez-Guerra et al., 2022). Neuropeptide-like molecules occur in all major animal clades with neurons, the Bilateria, Cnidaria and Ctenophora, and even in the neuron-less Placozoa and Porifera (Hayakawa et al., 2022; Koch and Grimmelikhuijzen, 2019; Nikitin, 2015; Sachkova et al., 2021; Yañez-Guerra et al., 2022). However, the deep relationships of animal neuropeptidergic systems, in particular between Bilateria and non-bilaterians, have remained elusive.

Most mature bioactive neuropeptides are 3-20 amino acids long and derive from longer proneuropeptide precursors through cleavage and other post-translational modifications. The same set of enzymes are involved in proneuropeptide processing across animals. These include prohormone convertases, which recognize dibasic cleavage sites flanking the active peptides, and peptidyl-glycine alpha-amidating monooxygenase (PAM), which convert C-terminal glycine residues to amide groups (Chufán et al., 2009; Seidah, 2011). Proneuropeptide precursors can contain a single peptide or multiple copies of identical or divergent peptide sequences separated by cleavage sites. The often repetitive structure and the presence of short active sequences interspersed with less-constrained interpeptide regions allow propeptide sequences to evolve relatively rapidly. Consequently, with increasing evolutionary distances it often becomes hard or impossible to recognize orthologous proneuropeptide sequences. Even within Bilateria, the orthology relationship of many neuropeptide families between protostomes and deuterostomes have only been recognized due to the orthology of their receptors (e.g., vertebrate orexin and insect allatotropin) (Alzugaray et al., 2019; Elphick et al., 2018; Mirabeau and Joly, 2013; Semmens et al., 2015).

Most neuropeptides signal through G protein-coupled receptors (GPCRs), which are larger seven-transmembrane proteins and show a slower evolutionary rate than proneuropeptide precursors. The cases where the evolution of both the proneuropeptides and their receptors could be reconstructed, revealed a strong pattern of coevolution between receptor and ligand (Elphick et al., 2018; Escudero Castelán et al., 2022; Grimmelikhuijzen and Hauser, 2012; Jékely, 2013; Mirabeau and Joly, 2013; Yañez-Guerra and Elphick, 2020). This coevolutionary pattern has been used to also trace the evolution of peptide families where the ligands are too divergent to retain phylogenetic signal. These analyses revealed more than 30 conserved peptidergic signaling systems across bilaterians (Elphick et al., 2018; Jékely, 2013; Mirabeau and Joly, 2013; Thiel et al., 2021; Yañez-Guerra et al., 2022).

The nervous system of cnidarians has long been known to be strongly peptidergic and proneuropeptides have been found across all major cnidarian clades (Koch et al., 2021; Koch and Grimmelikhuijzen, 2020, 2019). Most cnidarian neuropeptides are short amidated peptides, resembling those found in bilaterians. Genomic comparisons revealed three proneuropeptide families that were present in the cnidarian stem lineage. All three give rise to short amidated neuropeptides: GLWamides, GRFamides, and PRXamides (with “X” representing a variable amino acid residue) (Koch et al., 2021; Koch and Grimmelikhuijzen, 2020, 2019). Other neuropeptides are more specific to certain cnidarian groups such as QITRFamide and HIRamide to Hexacorallia or LRWamides to Anthozoa (Koch and Grimmelikhuijzen, 2020).

Peptides that are sufficiently similar to be considered orthologous between bilaterians and cnidarians are restricted to a few atypical neuropeptides. These include insulin-related peptides, glycoprotein-hormone-related peptides, trunk related peptides, nesfatin, and phoenixin (de Oliveira et al., 2019; Roch and Sherwood, 2014; Yañez-Guerra et al., 2022). The more common short amidated peptides, however, have no clear orthologs in bilaterians. Similarity is at most restricted to one or two C-terminal residues, such as between RFamide or Wamide peptides (Walker et al., 2009; Jékely, 2013; Williams, 2020). Sometimes this has been interpreted as evidence of common origin but there is no other evidence supporting potential orthologies for these cnidarian peptides and their bilaterian counterparts. Receptors for most cnidarian neuropeptides are still unknown, with two exceptions. These include receptors for Hydra vulgaris RFamide peptides and a receptor for a PRXamide maturation inducing hormone (MIH) in the hydrozoan Clytia hemisphaerica (Artigas et al., 2020; Assmann et al., 2014). Hydra RFamide peptides activate heterotrimeric peptide-gated ion channels belonging to the DEG/ENaC family and distantly related to bilaterian RFamide- and Wamide-gated ion channels (Dandamudi et al., 2022; Elkhatib et al., 2022; Gründer et al., 2022a). The Clytia MIH receptor is a class A GPCR, which shows a many-to-many ortholog relationship to a range of bilaterian neuropeptide families, several of which end in RFamide (Artigas et al., 2020).

Cnidarian genomes can encode a large number of class A GPCRs. The genome of the sea anemone Nematostella vectensis for example contains over 1000 GPCR genes (Krishnan and Schiöth, 2015). Earlier analyses suggested direct orthologous relationships between several cnidarian and bilaterian neuropeptide GPCRs such as orexin/allatotropin, somatostatin or neuropeptide Y receptors, beside others (Alzugaray et al., 2013; Anctil, 2009; Krishnan and Schiöth, 2015). However, these results were either only based on BLAST similarity, or limited phylogenetic analyses. More recent and comprehensive phylogenies suggest that cnidarian neuropeptide GPCRs are more closely related to each other than to bilaterian neuropeptide GPCRs and only show many-to-many, or few-to-many orthology with bilaterian receptors (Artigas et al., 2020; Hauser et al., 2022; Thiel et al., 2018). However, a comprehensive phylogenetic analysis of cnidarian GPCRs is still lacking. This, together with the paucity of experimentally characterized receptors, leaves our understanding of the evolution of eumetazoan peptidergic systems fragmentary.

Here, in a large-scale bioinformatic analysis, we map the global sequence diversity of metazoan class A GPCRs to identify neuropeptide GPCR candidates in cnidarians. We then use mass spectrometry and bioinformatics to compile a library of predicted N. vectensis neuropeptides. In a combinatorial pharmacological ligand-receptor assay, we test our peptide library against selected N. vectensis GPCRs and identify 31 neuropeptide receptors. By phylogenetic analysis, we reconstruct ancestral cnidarian GPCR A systems and their relationship to bilaterian systems. Finally, we map proneuropeptide and GPCR expression to a single-cell RNAseq dataset to analyze tissue-level peptidergic signaling in N. vectensis. Our results reveal extensive peptidergic networks in N. vectensis and support the independent expansion of neuropeptide signaling in Cnidaria and Bilateria.

Results and Discussion

Creation of a Nematostella neuropeptide library

To obtain a comprehensive library of endogenous neuropeptides in N. vectensis, we extended the list of known proneuropeptides (Hayakawa et al., 2019; Koch and Grimmelikhuijzen, 2020) by mass spectrometry and bioinformatic screening (Figure 1A). We first screened a N. vectensis transcriptome collection for sequences encoding a signal peptide. This predicted secretome was filtered with regular expressions to detect sequences with the repetitive dibasic cleavage sites (K and R in any combination) and amidation sites. In addition, we used standard BLAST searches using known cnidarian neuropeptide precursors. In parallel, we analyzed methanolic extracts from N. vectensis larval, juvenile and adult tissue by mass spectroscopy (LC-MS/MS) (Figure 1–figure supplement 1). The LC-MS/MS analysis confirmed the existence of peptides predicted from various precursor sequences (Supplementary file 1, 2), including precursors that produce identical peptides, such as the 2 proFLRNamide or proLRWamide1 precursors (Figure 1C) (Koch and Grimmelikhuijzen, 2020).

Identification of Nematostella vectensis neuropeptides.

(A) Pipeline to identify neuropeptides and their receptors and to reconstruct the evolution of cnidarian peptidergic signaling. (B) Peptide sequence logos of N-terminal and C-terminal peptide cleavage sites based on peptides detected by LC-MS/MS. Cleavage occurs at the dashed lines. (C) N. vectensis neuropeptide precursor schemes of peptides for which we identified a receptor, with sequence logos of the encoded peptide(s) on the left and length of precursor on the right. a = amide.

Mass spectrometry pipeline.

(A) Pipeline for mass spectrometry identification of neuropeptide candidates. (B) Example spectra of detected HIRamide peptides with different lengths that originate from the same peptide on the HIRamide precursor. The shown spectra are from the encircled HIRamide peptide copy in (C). (C) HIRamide precursor with different HIRamide peptides indicated as a blue line below the precursor sequence.

MS confirmed the occurrence of N-terminal peptide cleavage at dibasic KR-x, RR-x and KK-x sites, with “x” indicating the N-terminal amino acid of the resulting active peptide (Figure 1B, Supplementary file 1). Cleavage at such dibasic sites is typical for bilaterian precursors (Southey et al., 2008; Veenstra, 2000). In Nematostella, the cleavage site is often followed by a-Q residue as the most N-terminal amino acid of the peptide (Figure 1B). Cleavage at dibasic sites, however, does not occur in all precursors. Our MS data showed that some peptides are processed by cleavage N-terminal to a-Q residue without any basic residue but the cleavage site is instead often accompanied by an acidic (D or E) residue 1 or 2 positions N-terminally of the cleavage site (Figure 1B, Supplementary file 1). Another motif includes two proline residues in position 2 and 3 from the N-terminus of the peptide, and in some cases a single proline in either the 2nd or 3rd position (Figure 1B, Supplementary file 1). These proline-related cleavage sites are also often accompanied by an acidic residue flanking the peptide N-terminal of the cleavage site. Such non-dibasic cleavage sites have previously been proposed for Cnidaria (Hayakawa et al., 2022, 2019; Koch and Grimmelikhuijzen, 2019). Some precursors, such as the one for HIRamide, also showed alternative N-terminal cleavage of the same peptide copy, resulting in different versions of the same peptide (Figure 1–figure supplement 1, Supplementary file 1), indicating stepwise cleavage or controlled degradation of peptides. C-terminal peptide cleavage occurs at dibasic x-KR, x-RR, x-KK and in some cases at alternative monobasic Rxx-R sites, with a second basic amino acid 3 positions N-terminal of the cleavage site (Figure 1B, Supplementary file 1). This is similar to the C-terminal cleavage found in bilaterian propeptides (Southey et al., 2008; Veenstra, 2000). In many neuropeptides, we could also confirm peptide alpha-amidation by the conversion of a C-terminal glycine residue to an amide group.

Based on the MS data, we extended our regular expressions representing cleavage sites and re-screened the predicted secretome. With our combined approach, we could identify novel neuropeptide precursors, verify the processing of known neuropeptide precursors and refine cleavage site predictions. We used this information to prepare a list of 33 N. vectensis proneuropeptides, excluding potential isoforms of the same precursor but including 2 potential paralogs of the FLRnamide, LRWamide1, and QITRFamide precursors (Supplementary file 2). Our screen complements the list of known neuropeptides in N. vectensis (Hayakawa et al., 2022, 2019; Koch and Grimmelikhuijzen, 2020) with 15 new neuropeptide precursors. However, our list did not contain the recently identified GGYamide, GTEamide and IVLamide peptides (Hayakawa et al., 2022) or bursicon- and insulin-like peptides.

We inspected all precursors individually and predicted signal peptides, cleavage sites and amidation sites (Figure 1C, Supplementary file 2). Based on our cleavage-site predictions we then compiled a library of 64 synthetic Nematostella neuropeptides, including different versions of peptides from the same precursors and alternatively cleaved peptides that differ in the length of their N-terminus (Supplementary file 2 and Supplementary file 3).

Analysis of metazoan class A GPCRs and selection of N. vectensis neuropeptide-receptor candidates

To identify neuropeptide-GPCR candidates in N. vectensis, we focused on class A GPCRs representing the main type of neuropeptide receptors in bilaterians (Jékely, 2013; Mirabeau and Joly, 2013). We first aimed to get an overview of the diversity of class A GPCRs across metazoans and screened transcriptomes of nine cnidarian, six bilaterian, two placozoan, three ctenophore and five sponge species for these receptors (Supplementary file 5, 6). The number of GPCRs across Metazoa varies by species and seems not to correlate with phylogenetic affiliation (Figure 2A). From the combined N. vectensis transcriptomes (see Supplementary file 5) we initially detected a total of 1061 class A GPCR sequences. We analyzed these together with over 7000 GPCRs retrieved from the other species. As a reference, we used the experimentally confirmed and annotated human, Drosophila melanogaster, Caenorhabditis elegans, and Platynereis dumerilii sequences. We carried out a sequence-similarity-based clustering analysis of all sequences. This revealed only a few well-connected clusters of receptors conserved across all major metazoan lineages (Figure 2–figure supplement 1). These include the leucine-rich-repeat containing (lrrc) GPCRs, which are receptors for larger peptides such as bursicons and related glycoprotein hormones or for relaxins and related insulin-like peptides. Other well-connected GPCR clusters were more restricted to the individual phylogenetic groups. Mammalian olfactory receptors represent such a GPCR cluster with limited taxonomic breadth (Figure 2–figure supplement 1).

Cluster map of selected class A GPCRs.

(A) Number of class A GPCRs identified by HMMer search in the different investigated species. (B) Relationship of species used for cluster analysis in C. (C) Cluster analysis of major class A GPCR groups from cnidarian, bilaterian and placozoan species. Each dot represents a GPCR sequence with color-coding and symbols according to the phylogeny in B. Connecting lines between single sequences show similarity with P-values indicated in the top right. Cluster annotations are based on deorphanized bilaterian class A GPCRs. Abbreviations in C: ACh = acetylcholine, lrrc = leucine rich repeat containing, P2Y = purinergic P2Y receptor. Silhouette images in B were taken from phylopic.org.

Figure 2–source data 1. Raw cluster analysis CLANS file.

Clustermap of metazoan class A GPCRs.

Each dot represents a GPCR sequence with color-coding according to Species list on the left. Connecting lines between single sequences show similarity with P-values indicated in top right. Annotations are based on sequences of deorphanized bilaterian class A GPCRs. Abbreviations in C: ACh = acetylcholine, lrrc = leucine rich repeat containing, P2Y = purinergic P2Y receptor.

Figure 2–figure supplement 1–source data 1. Raw cluster analysis CLANS file.

For a more detailed analysis, we then reduced our sampled species to the cnidarian, the bilaterian with experimentally confirmed GPCRs and Petromyzon marinus, and the two placozoan species (Figure 2B). We re-run the clustering analysis and filtered out non-connected single sequences, species-specific cnidarian clusters and unconnected clusters only containing bilaterian or placozoan sequences. The two major clusters in our GPCR map contain bilaterian and cnidarian sequences, interspersed with only a few placozoan GPCRs (Figure 2C). In both major clusters, cnidarian and bilaterian sequences form separate subclusters, rather than cnidarian GPCRs being interspersed with bilaterian GPCRs. One of these clusters contains bilaterian GPCRs for low-molecular-weight neurotransmitters, including monoamines, trace amines, acetylcholine, melatonin and ATP. This cluster is weakly connected to the opsins. The second major cluster contains most bilaterian neuropeptide receptors (Figure 2C - neuropeptide II, III, gamma) as well as the bilaterian-specific chemokine, purinergic-P2Y and fatty acid receptors. There is a second cluster of bilaterian neuropeptide receptors weakly connected to this main cluster (Figure 2C - neuropeptide I), which has only few cnidarian sequences connected to it. Based on this cluster analysis, we selected 161 Nematostella GPCR sequences. We focused on full-length GPCRs that are associated with the bilaterian neuropeptide GPCR clusters. In addition, we chose candidates from non-connected clusters that are uncharacterized but ancestral to all cnidarians, except the leucine-rich repeat-containing GPCRs, which are known to be activated by insulin-related and glycoprotein-hormone-related peptides in bilaterians.

Deorphanization of 31 Nematostella neuropeptide receptors

To experimentally identify Nematostella neuropeptide GPCRs, we tested our selection of 161 N. vectensis GPCRs in a pharmacological assay for activation by peptides from our peptide library (Figure 3A, Supplementary file 3, 8). We expressed the candidate GPCRs in mammalian cells, together with a promiscuous Gqi protein and a luminescent G5A reporter (Figure 3A, Supplementary file 7). The peptides were separated into different mixes (Supplementary file 3), which were then tested on each GPCR (Supplementary file 4). Receptor-mix combinations that gave a positive signal were further resolved by testing the individual peptides of the mix to identify the activating ligand (Supplementary file 4). Peptides activating a receptor were then tested at different concentrations to record dose-response curves and determine EC50 values for each peptide-receptor pair (Supplementary file 9).

Dose-response curves of Nematostella neuropeptide GPCR pairs.

(A) Pharmacological assay and pipeline to identify peptide-GPCR pairs. (B) Dose-response curves of peptide-GPCR pairs with log peptide concentration plotted against normalized luminescence. GPCRs that are activated by the same peptide(s) are grouped together with peptide sequence shown above and peptide name highlighted in black. If several peptides activate the same receptor, peptide sequences are shown within the graph. Receptor identification number is encircled in the upper left of each curve, EC50 values are indicated in the lower right. (C) Histogram of EC50 values of peptide-GPCR pairs, showing only the lowest EC50 per GPCR. (D) Peptide-receptor pairings showing number of receptors activated by the different peptides. Connection strength indicates EC50 values.

Figure 3–source data 1. Tibble with all data points used to calculate the dose-response curves and EC50 values, in.csv format.

In this screen, we identified 31 N. vectensis GPCRs activated mostly in the nanomolar range by peptides from 14 different precursors (Figure 1C, 3B,C,D Supplementary file 10). The neuropeptides GLWL, PFHamide, VRHamide and QWamide each activate a different, single receptor (GLWL/R18 EC50 = 9E-7, PFHa/R36 EC50 = 1.2E-8, VRHa/R186 EC50 = 2.7E-10, QWa/R69 EC50 = 9E-7). Peptides ending in RPamide from the HRPamide and FRPamide precursors activate the same, single receptor R248. The HRPa peptide has a lower EC50 value, the FRPa peptide, however, has a lower threshold activation while reaching its maximum activation at higher concentrations with an overall more shallow curve slope (HRPa/R248 EC50 = 8.9E-9, HRPa/R248 slope = 1.75, FRPa/R248 EC50 = 2.3E-7, FRPa/R248 slope = 0.42) (Supplementary file 10).

The RFamide peptides with the sequence QGRFamide and QITRFamide are each encoded on separate precursors and despite their strong sequence similarities each activate separate receptors (QGRFa/R70 EC50 = 2.1E-6, QGRFa/R234 EC50 = 4.7E-9, QITRFa/R196 EC50 = 7.7E-11) although one of the two QGRFamide receptors is also activated by higher concentrations of QITRFamide (QITRFa/R70 EC50 = 1.7E-5). Given that R70 has EC50 values in a lower micromolar range for both RFamides, it is possible that this receptor has another unknown ligand that is more specific but shares some structural similarity to these RFamides. Two receptors are activated by HIRamide (R21 EC50 = 2.1E-9, R29 EC50 = 1.9E-7). The more sensitive R21, however, showed a strong base-activation at all tested concentrations, leading to a shifted EC50 for which we adjusted the minimum values (Figure 3–figure supplement 1). We also tested different copy versions of HIRamide peptides and most activated the two receptors in a similar concentration range (Figure 3–figure supplement 1).

Dose response curves with EC50 values and peptide precursor of different HIRamide peptides.

(A) Dose response curves of HIRamide peptide and GPR 21 + 29. Dose response curves show activation of receptors GPR21 and GPR29 with different HIRamide peptide versions from the same precursor. Sequences and EC50 values for different HIRamides are shown below the dose response curves. The graph in Figure 3 shows HIRamide3 as the peptide with the lowest EC50 values. GPR21 showed already at low concentrations a high base activation compared to the negative control, but did not show a clear increase in intensity at concentrations between 1e-13 and 1e-11. We therefore set the minimum to the value measured at a concentration of 1e-13 instead of the negative control and show a comparison of the corresponding graphs and EC50 values here. (B) HIRamide precursor. The N-terminal signal peptide of the precursors sequence is not highlighted, cleavage + amidation sites are highlighted in a darker gray. Tested peptides are underlined and coloured as shown underneath the dose response curves.

Figure 3–figure supplement 1 and 2–source data 1. Tibble with all data points used to calculate the Dose-response curves and EC50 values, in.csv format.

Four receptors are activated by FLRNamide (FLRNa/R26 EC50 = 7.4E-8, FLRNa/R187 EC50 = 7E-8, FLRNa/R197 EC50 = 1.1E-8, FLRNa/R230 EC50 = 4.7E-9), and one of these is also sensitive to higher concentrations of FYRVamide (FYRVa/R197 EC50 = 1.1E-6). However, we only tested the non-modified FLRNamide peptide and not the phenylacetyl-LRNamide as described to exist in sea anemones (Grimmelikhuijzen et al., 1990) which may be a better ligand.

The three LRWamide peptides QGLRWamide (LRWa1), QGGLRWamide (LRWa2) and GPPQGLRWamide (LRWa3), which are each encoded in a single copy on separate precursors (Figure 1C), cross-activate 4 different GPCRs. Each LRWamide seems to have one preferred receptor (LRWa1/R19 EC50 = 9.1E-8, LRWa2/R213 EC50 = 3.1E-8, LRWa3/R204 EC50 = 7.4E-10) plus a fourth GPCR (R193) which appears similarly sensitive to all LRWamide peptides (LRWa1/R193 EC50 = 6.6E-11, LRWa2/R193 EC50 = 2E-10, LRWa3/R193 EC50 = 6.9E-10). Receptor 19 and 204 have clearly shifted curves and lower EC50 values for their preferred peptide, while Receptor 213 has similar EC50 values for LRWa1 (EC50 = 5.4E-8) and LRWa2 (EC50 = 3.1E-8) but is more sensitive to LRWamide 2 at lower concentrations while reaching its maximum activation later.

Finally, the PRGamide peptide, which belongs to the ancestral cnidarian PRXamides, activates at least 13 different receptors (R28, R32, R198, R199, R200, R202, R210, R211, R219, R220, R221, R222, R223) in Nematostella with EC50 values between 1.6E-10 and 4.5E-8. We also tested R28 and R32 with longer versions of the PRGamide, as these were described elsewhere based on mass spectrometry data (Hayakawa et al., 2019). The higher EC50 values for these longer PRGamides (Figure 3–figure supplement 2) and the sequences on the precursor themselves (Supplementary file 2), however, suggest that these are not fully processed and the actual PRGamide is a tetrapeptide (GPRGamide and APRGamide) as previously suggested (Koch and Grimmelikhuijzen, 2020). The same is likely true for the longer QGRFamide version (QGRFGREDQGRFamide) (Hayakawa et al., 2019) which is also likely not fully processed as in this case the activation of the QGRFamide receptors in the initial screen was much lower for the mix that contained the longer version than for the mix containing the fully processed QGRFamide peptide (Supplementary file 4 - mix 4 vs. mix 6).

Dose response curves with EC50 values and peptide precursor of different PRGamide peptide versions for two of the PRGamide receptors.

(A) Dose response curves show activation of the receptors GPR28 and GPR32 with different PRGamide peptide versions from the same precursor. Sequences and EC50 values for different PRGamides are shown below the dose response curves. The graph in Figure 3 shows the shortest PRGamide (PRGa) as the peptide with the lowest EC50 values, which is likely the fully processed version. (B) PRGamide precursor. The N-terminal signal peptide of the precursor sequence is not highlighted, cleavage and amidation sites are highlighted in a darker gray. The tested peptides are coloured to match the color of the dose response curves and the GPRG peptide is underlined.

Figure 3–figure supplement 1 and 2–source data 1. Tibble with all data points used to calculate the Dose-response curves and EC50 values, in.csv format.

At least nine neuropeptide GPCR families are ancestral to cnidarians

To reconstruct the evolution of neuropeptide receptors in animals, we analyzed the phylogenetic relationships of the 31 deorphanized N. vectensis neuropeptide GPCRs to other cnidarian and bilaterian GPCRs. From the clustermap (Figure 2), we chose sequences with connection to the bilaterian and cnidarian neuropeptide GPCR clusters and calculated phylogenetic trees. In an initial analysis, we found that the bilaterian chemokine, purino, fatty acid and other related bilaterian-specific receptors are likely a diverged ingroup of the bilaterian neuropeptide gamma rhodopsin receptors (Figure 4–figure supplement 1,2, Supplementary file 11). Gamma rhodopsin receptors are specific to bilaterians and include somatostatin/allatostatin A, opioid/somatostatin/allatostatin C, kisspeptin and melanin-concentrating hormone receptors (Mirabeau and Joly, 2013; Thiel et al., 2021). We then deleted redundant group-specific expansions of loosely connected orphan clusters, the non-neuropeptide chemokine and related receptors, decreased the number of bilaterian species and calculated a detailed neuropeptide-GPCR phylogeny (Figure 4, Supplementary file 11).

Phylogeny of metazoan class A neuropeptide GPCRs.

(A) Phylogeny of species used in B. (B) Phylogeny of neuropeptide GPCRs with names of ligands. Branches are color coded according to A. Branches of deorphanized Nematostella GPCRs end in an asterisk. Alternating shades behind the tree branches highlight different monophyletic groups. Roman Numbers 1-3 and greek symbol gamma indicate approximate neuropeptide clusters shown in Figure 2. Left half circle of branch support indicates aBayes and the right half circle aLRT-SH-like support values. Detailed annotations in Supplementary file 11. (C) Table with number of receptors per group as highlighted in receptor phylogeny with a straight line indicating no receptor present. Two-letter abbreviations on top correspond to species in A. Abbreviations: a = amide, B = Bilateria, CCK = cholecystokinin, GnRH = gonadotropin releasing hormone, MIH = maturation-inducing hormone, Nm-U = neuromedin U, NpFF = neuropeptide FF, NpY/F = neuropeptide Y/neuropeptide F, P = Placozoa, PrP = prolactin releasing peptide, R.# = Nematostella GPCR number, sNpF = short neuropeptide F, t-FMRFa = trochozoan FMRFamide, TRH = thyrotropin releasing hormone.

Figure 4–source data 1. Raw sequences used for tree building,.fasta format.

Figure 4–source data 2. Aligned sequences used for tree building.

Figure 4–source data 3. Trimmed sequence alignment used for tree building.

Figure 4–source data 4. Tree file in nexus format.

Tree (FastTree) of neuropeptide GPCRs with bilaterian chemokine and related receptors.

Bilaterian sequences are shown in yellow, cnidarian sequences are shown in dark blue (Medusozoa) and magenta (Anthozoa), placozoan sequences are shown in light blue. Deorphanized Nematostella vectensis sequences are shown in green. Sequences were aligned using muscle, alignment was trimmed with the gappyout option of trimal, tree was calculated using FastTree. Abbreviations: a = amide, CCK = cholecystokinin, GnRH = gonadotropn releasing hormone, GPR19 = G protein-coupled receptor 19, L11 = elevenin, MCH = melanin concentrating hormone, Nm-U = neuromedin U, NpFF = neuropeptide FF, NpS = neuropeptide S, NpY/F = neuropeptide Y/neuropeptide F, PrP = prolactin releasing peptide, sNpF = short neuropeptide F, TRH = thyrotropin releasing hormone. Figure 4–figure supplement 1–source data 1. Raw sequences used for tree building,.fasta format.

Figure 4–figure supplement 1–source data 2. Aligned sequences used for tree building. Figure 4–figure supplement 1–source data 3. Trimmed sequence alignment used for tree building.

Figure 4–figure supplement 1–source data 4. Tree file in nexus format.

Tree (IQtree) of neuropeptide GPCRs with bilaterian chemokine and related receptors.

Bilaterian sequences are shown in yellow, cnidarian sequences are shown in dark blue (Medusozoa) and magenta (Anthozoa), placozoan sequences are shown in light blue. Deorphanized Nematostella vectensis sequences are shown in green. Sequences were aligned using mafft, alignment was trimmed with the gappyout option of trimal, tree was calculated using IQtree. Abbreviations: a = amide, CCK = cholecystokinin, GnRH = gonadotropin releasing hormone, GPR19 = G protein-coupled receptor 19, L11 = elevenin, MCH = melanin concentrating hormone, Nm-U = neuromedin U, NpB/W = neuropeptide B/neuropeptide W, NpFF = neuropeptide FF, NpS = neuropeptide S, NpY/F = neuropeptide Y/neuropeptide F, sNpF = short neuropeptide F.

Figure 4–figure supplement 2–source data 1. Raw sequences used for tree building,.fasta format.

Figure 4–figure supplement 2–source data 2. Aligned sequences used for tree building.

Figure 4–figure supplement 2–source data 3. Trimmed sequence alignment used for tree building.

Figure 4–figure supplement 2–source data 4. Tree file in nexus format.

The majority of known neuropeptide GPCRs are grouped into three main bilaterian clusters and two main cnidarian clusters (Figure 4B). Within the two cnidarian clusters there are at least nine neuropeptide GPCR families that are ancestral to Cnidaria, all of which are represented with clear anthozoan and medusozoan orthology groups. The deorphanized N. vectensis neuropeptide GPCRs belong to at least seven of these ancestral cnidarian families (Figure 4B, C).

One of the two main cnidarian clusters seems to have only expanded late within the Hexacorallia branch, with an especially high number of more than 100 GPCRs in N. vectensis (Figure 4B,C). The entire cluster is expanded in N. vectensis and Exaiptasia pallida, but neither expanded in Corallium rubrum, nor in medusozoan species. This cluster likely expanded from a single ancestral cnidarian receptor family as it only contains a single branch of medusozoan sequences. Alternatively, this cluster can be traced back to two peptidergic systems, with a loss of medusozoan sequences in one of them. This cluster contains the more promiscuous and less sensitive N. vectensis QGRFamide GPR70, the QWamide GPR69 and the PFHamide GPR36. The second major cnidarian neuropeptide GPCR cluster contains at least eight ancestral cnidarian receptor families, based on the presence of clear anthozoan and medusozoan orthology groups, with deorphanized N. vectensis receptors in six of them (Figure 4B,C). The QITRFamide GPR196 belongs to a small group of hexacorallian receptors with no clear orthologs in C. rubrum or any medusozoan species, which might either represent a separate ancestral group with no medusozoan representative sequences or a strongly diverged Hexacorallia subcluster. This is in accordance with the absence of QITRFamide peptides in Octocorallia and Medusozoa (Koch and Grimmelikhuijzen, 2020). The QGRFamide GPR234 belongs to an ancestral cnidarian GPCR family, which is in accordance with the existence of QGRFamide or GRFamide peptides across Cnidaria (Koch and Grimmelikhuijzen, 2020, 2019). This group is slightly expanded in the medusozoans Clytia hemisphaerica and Hydra vulgaris. The entire QGRFamide family seems related to the LRWa GPR19 containing group, which has otherwise no direct orthologous sequences in medusozoans. The GLWL peptide GPR18 belongs to an ancestral cnidarian family that shows a slight expansion in N. vectensis. The receptors for the peptides LRWamide, FLRNamide and HIRamide are closely related as part of an anthozoan/hexacorallian expansion, curiously twice in unrelated families. Each of the families is ancestral to cnidarians and one expanded in anthozoans and additionally contains the receptors for VRHamide and RPamides, while the other group only showed several sequences in the two analyzed Hexacorallia species N. vectensis and E. pallida. Both families contain receptors for the three peptides, LRWamide, FLRNamide and HIRamide, and both families only contain medusozoan sequences from R. esculentum and C. cruxmelitensis. The 13 PRGamide receptors of N. vectensis belong to two separate ancestral cnidarian families, both of which are expanded in anthozoans. One PRGamide family independently also expanded in Hydra vulgaris. The other PRGamide family contains the Clytia MIH receptor which is activated by Clytia PRXamides (Artigas et al., 2020), confirming a peptide-receptor coevolution in cnidarians for the ancestral PRXamides (Koch and Grimmelikhuijzen, 2019) and their receptor(s).

Cnidarian and bilaterian neuropeptide GPCR systems expanded after the cnidarian-bilaterian split from a few ancestral systems

Our phylogenetic analyses divided the cnidarian and the bilaterian neuropeptide receptors into very few major clusters, each containing either bilaterian or cnidarian sequences (Figure 4B). The relationship of these major clusters to each other is not well resolved. However, the general clustering (Figure 2C) in combination with the phylogenetic analysis (Figure 4B) shows that these groups represent many-to-many bilaterian-cnidarian orthology groups. Most of the bilaterian and cnidarian class-A neuropeptide GPCRs therefore diversified after the bilaterian-cnidarian split. The only family with an apparently consistent one-to-one orthology of an ancestral cnidarian and bilaterian branch of receptors is that of the bilaterian orphan receptor GPR19.

In addition there are a few branches with unclear phylogeny or restricted taxonomic breadth. A small group of anthozoan orphan receptors without medusozoan representatives shows affinity to the expanded cluster of bilaterian oxytocin/vasopressin, GnRH, achatin and neuropeptide S/CCAP receptors, suggesting a one-to-many orthology. This affinity is consistent in different trees (compare Figure 4 and Figure 4–figure supplement 1,2, or Supplementary file 11). Other orthology groups between Cnidaria and Bilateria are less clear and the potential branches in question are not stable between the different phylogenies. The bilaterian SIFamide GPCR grouped together with an orphan cnidarian branch, but varied in its position in other analyses (compare Figure 4 and Figure 4–figure supplement 1 and 2, or Supplementary file 11). This instability is also obvious due to the fact that the SIFamide GPCR is the protostome ortholog of the deuterostome NpFF GPCRs, which would have been expected to group together.

Accordingly, the NpFF receptor was also unstable in our analyses. The NpFF and SIFamide receptor group is usually well supported, but often shows long basal branches and these two together built in previous analyses a more separated branch with unstable relationship to other bilaterian receptor groups (Mirabeau and Joly, 2013; Thiel et al., 2021, 2018). The bilaterian QRFPR branch has a likewise unstable relationship to other bilaterian neuropeptide GPCRs in previous analyses but showed affinity to the SIFamide/NpFF GPCRs in some analyses (Jékely, 2013; Thiel et al., 2018). Accordingly, the affinity of some N. vectensis sequences to the QRFPR branch is also not stable in our different analyses (compare Figure 4 and Figure 4– figure supplement 1 and 2; Supplementary file 11). Two ancestral cnidarian branches grouped together with the Drosophila trissin receptor, but also this is an unstable grouping not present in our supplementary analyses (Figure 4–figure supplement 1 and 2). The trissin receptor is a generally peculiar bilaterian protostome receptor as no orthologous deuterostome sequences are known (Elphick et al., 2018; Mirabeau and Joly, 2013; Thiel et al., 2021). The orexin/allatotropin receptors showed no direct orthology to cnidarian receptors, but appear as a sister group to a branch with orphan bilaterian and placozoan GPCRs. Together, except for the GPR19 group, we could not find cnidarian receptors that show a consistent 1:1 orthology relationship to specific bilaterian receptors such as for example proposed for orexin/allatotropin, somatostatin, neuropeptide Y or tachykinin receptors (Alzugaray et al., 2019; Anctil, 2009; Krishnan and Schiöth, 2015). This is in accordance with other analyses that used a wider array of bilaterian neuropeptide GPCRs when comparing them to cnidarian GPCR sequences (Artigas et al., 2020; Hauser et al., 2022; Thiel et al., 2018) and found rather many-to-many orthologs, if any at all.

Cell-type-specific expression of neuropeptides and GPCRs and the peptidergic connectome of Nematostella

To analyze tissue-level peptidergic signaling in N. vectensis, we mapped the expression of neuropeptide precursors and the newly deorphanized neuropeptide GPCRs to a single-cell RNAseq dataset (Steger et al., 2022). The single-cell data are split into two sets, one consisting of pooled stages spanning 18 hours post-fertilization to 16-day-old primary polyp (developmental set) and the second set consisting of adult tissues only, similar to (Cole et al., 2023). The expression of GPCRs was in many cases low and only a small percentage of cells within a given cell type show receptor expression. We could not detect expression of GLWL receptor R18 and HIRamide receptor R29, consistent with the generally low expression of GPCRs in animals (Fredriksson and Schiöth, 2005; Regard et al., 2008; Soave et al., 2021; Sriram et al., 2019). In contrast, neuropeptide precursors are generally highly expressed in neurons (Smith et al., 2019) and we could detect all of them, except for the PFHamide that could not be mapped and was only present as a partial sequence in our combined transcriptome (Figure 1C). Most neuropeptide precursors show restricted expression in neuroglandular cells (Figure 5B), with the exception of the phoenixin and GLWL precursors (Figure 5–figure supplement 1). The broad expression of phoenixin suggests other functions for this molecule outside neuronal signaling, for example in mitochondrial regulation (Dennerlein et al., 2015; Yañez-Guerra et al., 2022).

Tissue-specific expression.

Dotplot for tissue-specific expression of peptide precursors and GPCRs. Red dots indicate expression in the developmental dataset, blue dots indicate expression in the adult dataset. Abbreviations: a = amide, e = embryonic, ect = ectoderm, endomes = endomesoderm, gland = glandular, muc = mucous, musc = muscle, neurogland = neuroglandular, PGC = primary germ cells, pharyng = pharyngeal, pSC = putative stem cells, R = receptor (GPCR), retrac = retractor.

Tissue-specific expression of proneuropeptides and neuropeptide GPCRs.

Dotplot for tissue-specific expression of proneuropeptides and GPCRs. Proneuropeptides without a known receptor are also included. Red dots indicate expression in the developmental dataset, blue dots indicate expression in the adult dataset.

Individual GPCRs were often restricted to a single or very few tissue types (Figure 5B for tissue type overview and Supplementary file 12 for cell-type-specific resolution). For peptides with multiple receptors, we often found distinct patterns of receptor expression. Individual PRGamide receptors are for example restricted to the retractor muscle (R221, R222), the gastrodermis (R211), embryonic putative stem cells (pSC) (R210) or a combination of neuroglandular cells and adult cnidocytes (R202) (Figure 5B). An exception is the PRGamide receptor R200, which shows expression in a wide array of tissues, including adult pSC, primary germ cells (PGCs), neuronal cells, immune cells, and embryonic ectodermal cells.

The different LRWamide receptors also signal to different tissue types. The receptor R204, which is most sensitive to LRWamide3, is strongest expressed in embryonic endomesodermal cells and the pharyngeal ectoderm, the receptor R19, which is most sensitive to LRWamide1 is found in neuroglandular cells, the receptor R213, which showed highest sensitivity to LRWa2 is found in neuroglandular and glandular mucousal cells and the receptor R193, which is similarly sensitive to all LRWamides is only present in glandular mucousal cells. The two receptors of the ancestral QGRFamide are also differentially expressed, with the receptor R70 only present in adult neuroglandular cells, while the highly specific R234 is additionally expressed in developmental and adult glandular mucousal cells and in some developmental retractor muscle cells.

Within the neuroglandular subset, each neuronal cell type expresses a unique combination of neuropeptide precursors and GPCRs (Figure 5–figure supplement 2 and 3; Supplementary file 12). Most peptide precursors are thereby restricted to only a few cell types. PRGamide for example, is restricted to two types of neuroglandular cells: the larval apical organ Nl(+)ao.PRGa cells and the Ne(-)PRDM14d cells, which persist in adults (Figure 5–figure supplement 2,3). This restricted expression of the PRGamide precursor is also in accordance with previously published in situ hybridizations that show a restricted expression in the larval apical organ (Gilbert et al., 2022). Many neuronal cell types only express a low number of different neuropeptide precursors, with few exceptions such as the mentioned Nl(+)ao.PRGa cells or the Nl(+)rxraB and N(+)GFI1B.prdm cells, which express between 4-8 different types of neuropeptide precursors, depending on the stage. The larval Nl(+)ao.PRGa cells express the PRGamide, VRHamide, OITRFamide, RPamide, LRE peptide, GLWL peptide and the phoenixin precursor, while only expressing two receptors: the PRGamide receptor R.200 with a lower average expression and the VRHamide receptor R186. The expression of both the PRGamide and VRHamide peptide precursors and receptors may indicate autocrine regulation in these cells. The second PRGamide-positive cell type Ne(-)PRDM14d that persists in adults, also shows in both datasets an expression of the PRGamide precursors and a different PRGamide receptor R.221, which is only found in few cell types and similarly may mediate autocrine feedback.

Expression of neuropeptide precursors and GPCRs in neuroglandular cell types in the developmental dataset.

Dot size indicates percentage of cells expressing the corresponding gene, color intensity indicates average expression.

Expression of neuropeptide precursors and GPCRs in neuroglandular cell types in the adult dataset.

Dot size indicates percentage of cells expressing the corresponding gene, color intensity indicates average expression.

To determine the global organization of peptidergic signaling networks or the ‘peptidergic connectome’ (Bentley et al., 2016; Deng et al., 2019; Smith et al., 2019; Williams et al., 2017) in Nematostella, we constructed a multi-layered network for both the developmental and the adult dataset (Figure 6, Figure 6–figure supplement 1). In these networks, nodes represent cell/tissue types and links are defined from peptide-expressing cells to receptor-expressing cells taking into account expression levels and the EC50 values. Different peptide-receptor pairs (represented by different colors in Figure 6, Figure 6–figure supplement 1 B-D, F-L) form distinct layers in this multi-layer connectome. By modularity analysis, we subdivided the networks into three (developmental) or five (adult) modules, each dominated by peptides with multiple receptors: PRGamide, QGRFamide, FLRNamide and LRWamide (Figure 6–figure supplement 1 A,E). The cells at the center of these modules that are most strongly involved in peptidergic signaling are the Ne(+)pit1, Ns(+)otp, Nl(+)ao.PRGa, Nl(+)rxraB, and Ne(-)PRDM14d cells in the developmental dataset and the Ne(+)pit1, Ns(+)otp, Ne(+)otx.early and Ne(-)PRDM14d cells in the adult dataset.

Multi-layer peptidergic connectomes in Nematostella.

Peptidergic networks in the (A) developmental and (B) adult subset. Nodes represent cell types, connections represent potential peptidergic signaling from neuropeptide-expressing cells to cells expressing one or more of the receptors for that neuropeptide. Colors represent different peptide-receptor signal channels (the different layers in the multi-layer connectome).

Figure6–source data 1. Graph file of the multilayered peptidergic connectome in the developmental subset. A serialized binary file in tbl_graph format, to be analyzed in R.

Figure 6–source data 2. Graph file of the multilayered peptidergic connectome in the developmental subset. In Gephi.gexf format, to be analyzed in Gephi.

Figure 6–source data 3. Graph file of the multilayered peptidergic connectome in the adult subset. A serialized binary file in tbl_graph format, to be analyzed in R.

Figure 6–source data 4. Graph file of the multilayered peptidergic connectome in the adult subset. In Gephi.gexf format, to be analyzed in Gephi.

Multi-layer peptidergic connectomes in Nematostella.

(A) Network of all peptide-receptor pairs for the developmental subset, coloured by network Leiden module. (B-D) Networks of LRWa (B), FLRNa (C) and PRGa (D) for the developmental subset. (E) Network of all peptide-receptor pairs for the adult subset, coloured by network Leiden module. (F-L) Networks of LRWa (F), FLRNa (G) PRGa (H), QGRFa (I), HIRa (J), VRHa (K) and QWa (L) for the adult subset. Nodes represent cell types, connections represent potential peptidergic signaling from neuropeptide-expressing cells to cells expressing one or more of the receptors for that neuropeptide. For peptides with more receptors, different colors represent different peptide-receptor signal channels (the different layers in the multi-layer connectome).

Discussion

We deorphanized 31 neuropeptide GPCRs of the sea anemone Nematostella vectensis and reconstructed their evolution and relationship to other cnidarian orphan GPCRs and bilaterian neuropeptide GPCRs. Our phylogeny suggests that the identified neuropeptide GPCRs belong to two major GPCR clades that expanded independently of each other. One clade diversified in stem Cnidaria into eight receptor families that are all present across medusozoans and anthozoans. The other clade diversified later within the anthozoans, likely from a single ancestral cnidarian receptor. The expansion of GPCR systems is a general feature of GPCR evolution and happened independently in different lineages and at different times during animal evolution (Beets et al., 2022; Mirabeau and Joly, 2013).

The cnidarian neuropeptide receptors we identified here show many-to-many orthology relationship to bilaterian neuropeptide GPCRs. We did not identify any direct receptor orthologs between cnidarians and bilaterians, indicating extensive parallel expansion of neuropeptide signaling in the two clades. Previously, some neuropeptides have been hypothesized to be orthologous across cnidarians and bilaterians, including neuropeptides ending in an RFamide (Jékely, 2013; Walker et al., 2009), which is a frequent motif (Elphick and Mirabeau, 2014; Jékely, 2013; Walker et al., 2009). We identified two N. vectensis RFamide receptors, one for the ancestral cnidarian neuropeptide QGRFamide (Koch et al., 2021; Koch and Grimmelikhuijzen, 2019) and one for the hexacorallian QITRFamide. These two N. vectensis receptors belong to either of the two major clades of cnidarian receptors, suggesting that new RFamide receptors and/or peptides can evolve independently. Likewise, within bilaterians there are many RFamides across protostomes and deuterostomes that are not orthologous to each other (Elphick and Mirabeau, 2014; Thiel et al., 2021). The cnidarian RFamide receptors are also more closely related to other cnidarian neuropeptide receptors than to bilaterian RFamide receptors, arguing against the orthology of cnidarian and bilaterian RFamides. C-terminal aromatic amino acids are found in various neuropeptides and also many monoamine receptors are activated by derivatives of aromatic amino acids (Bauknecht and Jékely, 2017; Elphick and Mirabeau, 2014). Aromatic amino acids might have been early ligands in the evolution of monoamine neurotransmission and ancestrally present as a common structural feature at the C-termini of neuropeptides activating peptide receptors.

Our phylogenetic analyses and GPCR resource will be useful to directly predict the ligand of neuropeptide receptors in other cnidarian species. We expect this because of the long-term coevolution of peptide-receptor pairs, that has been extensively documented in bilaterians. The orthology of six of the identified anthozoan N. vectensis PRGamide receptors to the PRXamide receptor of the hydrozoan C. hemisphaerica is the first example of a similar long-term ligand-receptor association across cnidarians. Receptors of ancient peptides such as PRXamides or GRFamide are therefore expected to be orthologous even in distantly related cnidarians. Other Hexacorallia-or Anthozoa-specific peptides such as HIRamide, FLRNamide and LRWamide (Koch and Grimmelikhuijzen, 2020) can be tested on directly orthologous receptors. Phylogenetic analyses combined with targeted deorphanization experiments of predicted orthologs could thus lead to the rapid characterisation of new receptors in other cnidarians. Our large-scale GPCR-peptide resource also allowed us to analyze tissue-level peptidergic networks in N. vectensis. With high-resolution single-cell resources becoming available in a larger number of species, it will be possible to predict the cellular targets of neuropeptide signals with increasing precision based on receptor expression. The unique combinations of proneuropeptides and neuropeptide receptors in the different neuronal cell types can also be used as markers to identify and characterize neuronal cell types. The uniquely specific combinatorial expression of proneuropeptides as neuronal markers parallels the situation found in bilaterian nervous systems, where proneuropeptides provide fingerprint-like identities to different neuronal cell types (Smith et al., 2019; Taylor et al., 2021; Williams et al., 2017). Finally, our work also opens up new avenues in experimental neuroscience in cnidarians (Bosch et al., 2017). With readily available genetic manipulation techniques in N. vectensis and other species (Ikmi et al., 2014; Nakanishi and Martindale, 2018; Paix et al., 2023) (Artigas et al., 2020; Sanders et al., 2018; Wittlieb et al., 2006), the identified neuropeptide-receptor interactions will enable genetic manipulations of both ligand and receptor(s), to reveal the biological functions of peptidergic signaling.

Overall, we identified receptors for peptides from nearly half of the Nematostella neuropeptide precursors, including the receptors of the ancient PRXamide and QGRFamide peptides. Future studies aiming at finding the remaining receptors could focus on GPCRs with no sequence similarity to known bilaterian neuropeptide GPCRs, including monoamine-related or leucine-rich-repeat receptors, or other types of receptors like DEG/ENaC-related ion channels (Gründer et al., 2022b; Mirabeau and Joly, 2013).

Supporting information